---
title: "CourseProject1"
author: "doyougnu"
date: "June 16, 2015"
output: html_document
---

##Synopsis
This report analyzes data from the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. The goal of this report is twofold: 1.
To determine which weather event, in the given database, is most harmful to 
human health. 2. Of the recorded weather events, which events are correlated to 
have the greatest amount of economic consequences. Economic consequences is used
vaguely here, this convievably could be a benefit, or a hinderance. Given the 
dataset this author chose to focus on economic damage, in place of economic
consequences. It was found that....

##Data Preprocessing
The following environment was used to process, and analyze the date:

```{r environment, echo=TRUE}
Sys.info()
R.Version()
```

The data is available [here][1], and comes in bzip2 form. To load this into R
we perform the following operations:

```{r dataPreprocessing, echo=TRUE, cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "FStormData.csv.bz2", method = "curl")
df <- read.csv(bzfile("FStormData.csv"), header = TRUE, sep = ",")
```

Now lets have a look at the first few lines and how R is interpreting the data:

```{r, quickLook, echo=TRUE}
str(df)
```

This analysis is focused on the EVTYPE (Event Type) variable, as such it must be
cleaned in order to be useful. First of all we cannot have "THUNDERSTORM" and 
"TSTM" mean the same thing in our analysis. Also, white spaces and special 
characters must be removed, and any other "EVTYPE" that does not make sense for
this analysis:

```{r EVTYPEclean, echo=TRUE}
#remove white spaces
df$EVTYPE <- gsub("^\\s+|\\s+$", "", df$EVTYPE)
df$EVTYPE <- tolower(df$EVTYPE)

#remove summaries, monthly/weekly summaries
patternsToRemove <- c("monthly", "?")
greplString <- paste(patternsToRemove, collapse = "|")
df <- df[(grepl(greplString, df$EVTYPE)),]
df <- df[(!grepl("summary", df$EVTYPE)),]
dim(df)
```


<!---Links--->
[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2