---
title: "CourseProject2"
author: "doyougnu"
date: "June 16, 2015"
output: html_document
---

##Synopsis
This report analyzes data from the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. The goal of this report is twofold: 1.
To determine which weather event, in the given database, is most harmful to 
human health. 2. Of the recorded weather events, which events are correlated to 
have the greatest amount of economic consequences. Economic consequences is used
vaguely here, as one could analyze the oppurtunity cost of lost lives or crops,
or one could merely ignore consequences of a lost. Given the 
dataset this author chose to focus on economic damage, in place of economic
consequences. It was found that the tornado is the most costly weather event in
terms of human health. The most economically damaging weather event was found 
the be a flood. In sum, both of these results imply that the dataset should 
be analyzed in  terms of geographic data, because floods and tornadoes are 
location specific.

##Data Preprocessing
The following environment was used to process, and analyze the date:

```{r environment, echo=TRUE}
Sys.info()
R.Version()
```

The data is available [here][1], and comes in bzip2 form. To load this into R
we perform the following operations:

```{r dataPreprocessing, echo=TRUE, cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "FStormData.csv.bz2", method = "curl")
df.raw <- read.csv(bzfile("FStormData.csv"), header = TRUE, sep = ",")
```

Now lets have a look at the first few lines to see how R is interpreting 
the data, and the subset the dataframe:

```{r, quickLook, echo=TRUE}
str(df.raw)
df <- subset(df.raw, select = c("PROPDMGEXP", "PROPDMG", "CROPDMGEXP", "CROPDMG",
                            "BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES"))
```

This analysis is focused on the EVTYPE (Event Type) variable, as such it must be
cleaned in order to be useful. First of all we cannot have "THUNDERSTORM" and 
"TSTM" mean the same thing in our analysis. Also, white spaces and special 
characters must be removed, and any other "EVTYPE" that does not make sense for
this analysis:

```{r EVTYPEclean, echo=TRUE, cache=TRUE}
#remove special characters
library(dplyr)
df$EVTYPE <- gsub("[[:punct:]]", "", df$EVTYPE)
df$EVTYPE <- tolower(df$EVTYPE)

#remove summaries, monthly/weekly summaries
patternsToRemove <- c("monthly", "?")
greplString <- paste(patternsToRemove, collapse = "|")
df <- df[(grepl(greplString, df$EVTYPE)),]

#remove everything that has summary in it
df <- df[(!grepl("summary", df$EVTYPE)),]

#turn TSTM into Thunderstorm
tstorm <- c("thunderstorms", "thunderstormw", "tstm", "thunderstorminds", "thunderstorminds", "thunderstrom", "thundertorm", "thundertsorm", "thundestorm", "thunerstorm", "thunderstorminds")
tstormGsub <- paste(tstorm, collapse = "|")
df$EVTYPE <- gsub(tstormGsub, "thunderstorm", df$EVTYPE)
df$EVTYPE <- factor(df$EVTYPE)
```

Now clean the economic data by converting the PROPDMGEXP and CROPDMGEXP variables
to scientific notation, concatenating with PROPDMG/CROPDMG, and generating the
final DMG variables

```{r econDataClean, echo=TRUE, message=FALSE, cache=TRUE, warning=FALSE}
fixDMG <- function(x){
#function takes a list, loops through the list converting values to scientific
#notation, converts list class to character to anticipate future paste call
    x <- toupper(x)
    x <- gsub("[[:punct:]]", "", x)
    x <- lapply(toupper(x), function(x) 
        ifelse(x == "K", "000", 
               ifelse(x =="M", "000000", 
                      ifelse(x == "B", "000000000", x))))
    
    x <- as.character(unlist(x))
    x
}
df$CROPDMGEXP <- fixDMG(df$CROPDMGEXP)
df$PROPDMGEXP <- fixDMG(df$PROPDMGEXP)

#convert and concatenate damage variables
df$PROPDMG <- as.character(df$PROPDMG)
df$CROPDMG <- as.character(df$CROPDMG)
df$propertyDamage <- as.numeric(paste(df$PROPDMG, df$PROPDMGEXP, sep = ""))
df$cropDamage <- as.numeric(paste(df$CROPDMG, df$CROPDMGEXP, sep = ""))
```

##Model Definition
To asses the impact of Climate Events to human health and the Economy I will be
using a simple sum model. Thus INJURIES and FATALITIES will be summed across the
all time periods in the data frame. The same type of model will be used to 
analyze the economic damage as well. I prefer to use the [dplyr][2] package for
operations like this:

```{r modelDef, echo = TRUE, message=FALSE}
#aggregate for human impact analysis
library(dplyr)

populationHealth <- df %>%
                       group_by(EVTYPE) %>%
                            summarise(humanInjuries = 
                                        sum(INJURIES, na.rm = TRUE),
                                      humanDeaths = 
                                        sum(FATALITIES, na.rm = TRUE), 
                                      total =
                                        sum(INJURIES,
                                            FATALITIES, na.rm = TRUE)) %>%
                                arrange(desc(total))

#aggregate for economic impact
econHealth <- df %>%
                group_by(EVTYPE) %>%
                    summarise(propertyImpact = 
                                  sum(propertyDamage, na.rm = TRUE),
                              cropImpact =
                                  sum(cropDamage, na.rm = TRUE),
                              total = 
                                sum(cropDamage,
                                    propertyDamage, na.rm = TRUE)) %>%
                        arrange(desc(total))
```


##Results
Here is the top ten damaging events by Economic damage and impact to human health
sorted by total economic and human impact:

```{r topTens, echo = TRUE}
populationHealth[1:15,]
econHealth[1:15,]
```

I would expect earthquakes to be at the top of the list because given the model
the event type which maximizes injuries and deaths will also maximize the 
"total impact". Likewise for the economic damage, however earthquakes are not
defined as a "weather event" in this dataset.

Here are some plots to describe the most costly weather events in terms of 
economic and human impact:

```{r plots, echo=TRUE, message=FALSE, warning=FALSE}
#load ggplot2
library(ggplot2)

#subset data for plots
populationHealth.sub <- populationHealth[1:15, 1:3]
econHealth.sub <- econHealth[1:15, 1:3]

#melt data to long tidy form
library(tidyr)
populationHealth.sub.tidy <- populationHealth.sub %>%
    gather(impactType, impact, 2:3) %>%
        arrange(impactType, desc(impact))

econHealth.sub.tidy <- econHealth.sub %>%
        gather(impactType, impact, 2:3) %>%
            arrange(impactType, desc(impact))

#plots
ggplot(populationHealth.sub.tidy, aes(x = factor(EVTYPE, levels = EVTYPE), 
                                      y = impact,
                                      fill = impactType)) + 
    geom_bar(stat = "identity") +
    coord_flip() +
    ggtitle("Impact of Weather Events on Human Health") +
    labs(y = "Human Impact by Fatalities and Injuries", x = "Event Type")

ggplot(econHealth.sub.tidy, aes(x = factor(EVTYPE, levels = EVTYPE),
                                y = impact/100000000,
                                fill = impactType)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    ggtitle("Economic Impact by Weather Event") +
    labs(y = "Economic Impact in Billions of Dollars", x = "Event Type")
```

Taking the plots in order: The most costly weather event in terms of human
wellbeing (defined as the sum of injuries and deaths) is the occurance of a
tornado. Tornado's dominate the dataset by an order of magnitude in injury
count. As this model treats an injury equal to a death this assymetry causes
the model to conclude the tornado as the most costly weather event to human
wellbeing. In the future this model could be revised, or merely reduced to a 
peerly economic analysis; while some might find equating a life to a price value
the oppurtunity cost of a death is substantial and of economic importance, so 
that analysis may have some merit.  

Turning to the economic analysis, the most costly weather event is a flood. For
similary reasons to the tornado, the flood dominates the model by maximizing the
property impact variable. The results of flood and tornados dominating the model
leads me to think that the dataset may be biased towards the southwest. I would
certainly expect it to be biased towards population centers - as there is no
health nor economic impact without people. Future analyses may try to analyze
this dataset in relation to geographic or socioeconomic data; or for that matter
reduce to the cropImpact veriable to a property Impact metric, as a crop is a
form of property.


<!---Links--->
[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
[2]: http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html